---
title: "Getting Started with accessMap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with accessMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 8,
  fig.height = 6,
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(accessMap)
```

# Introduction

The `accessMap` package provides tools for calculating and visualizing spatial accessibility to facilities such as health centers, schools, or emergency services. It uses cost-distance analysis based on travel time and generates publication-ready maps and summary statistics.

# Basic Usage

## Your First Accessibility Analysis

Let's start with a simple analysis using the default Kenya data:

```{r basic-example}
# Define some health facility coordinates in Kenya
health_facilities <- matrix(c(
  36.8219, -1.2921,
  39.6682, -4.0435,
  35.2834, 0.5143,
  34.7519, 0.0236 
), ncol = 2, byrow = TRUE)

# Add column names
colnames(health_facilities) <- c("lng", "lat")

# Run accessibility analysis
result <- accessibility(
  points = health_facilities,
  label = "Health Facilities",
  name = "kenya_health_example",
  save_files = FALSE  # Don't save files in this example
)

# View the summary statistics
print(result$national_summary)
```

The results show how Kenya's population is distributed across different travel time bands to the nearest health facility.

## Visualizing Results

The package creates three types of visualizations:

```{r visualizations, eval = FALSE}
# Combined map and summary chart
plot(result$combined_plot)

# Just the accessibility map
plot(result$map_plot)

# Just the population distribution chart
plot(result$bar_plot)
```

# Working with Built-in Data

The package includes several built-in datasets for Kenya that you can access directly:

```{r builtin-data, eval = FALSE}
# Access Kenya administrative boundaries
plot(accessMap::adm0)  # Country boundary
plot(accessMap::adm1)  # County boundaries
plot(accessMap::adm2)  # Sub-county boundaries
plot(accessMap::adm3)  # Ward boundaries

# View population distribution
plot(accessMap::population_raster)

# Load all default data at once
kenya_data <- load_default_data()
names(kenya_data) # Note, subcounty and ward shapefiles are not included in this function
```

# Custom Travel Time Categories

## Default Categories

By default, the package uses a 5-category system:

```{r default-categories}
# View default categories
cats <- default_categories()
print(cats$labels)
print(cats$colors)
```

## Creating Custom Categories

You can create custom categories for specific use cases:

```{r custom-categories}
# Emergency services (shorter time bands)
emergency_cats <- create_categories(
  breaks = c(0, 5, 10, 15, 30, Inf),
  labels = c("Critical", "Urgent", "Priority", "Standard", "Remote"),
  palette = "Reds"
)

# Auto-generate labels from breaks
rural_cats <- create_categories(
  breaks = c(0, 60, 120, 180, Inf) 
)

print(rural_cats$labels)
```

## Using Custom Categories

```{r custom-analysis, eval = FALSE}
# Use emergency categories for ambulance analysis
ambulance_result <- accessibility(
  points = ambulance_stations,
  categories = emergency_cats,
  label = "Ambulance Stations",
  save_files = FALSE
)
```

# Administrative-Level Analysis

## County-Level Summaries

Generate summaries for administrative units like counties:

```{r admin-analysis, eval = FALSE}
# First, get your accessibility data
access_data <- result$processed_data

# Generate county-level summary
county_summary <- generate_admin_summary(
  data = access_data,
  admin_sf = accessMap::adm1,
  admin_name_cols = "NAME_1"
)

# View results
head(county_summary)

```

# Advanced Visualization

## Flexible Point Styling

Create sophisticated maps with different point styles:

```{r advanced-styling, eval = FALSE}
# Create facility data with attributes
facilities_df <- data.frame(
  lng = c(36.8, 37.0, 36.9, 35.3),
  lat = c(-1.3, -1.5, -1.4, 0.5),
  type = c("Hospital", "Clinic", "Hospital", "Health Center"),
  ownership = c("Public", "Private", "Public", "Public"),
  capacity = c(200, 50, 150, 75)
)

# Create map with color by type, shape by ownership, size by capacity
advanced_map <- create_map(
  data = result$processed_data,
  points = facilities_df,
  adm0 = accessMap::adm0,
  adm1 = accessMap::adm1,
  categories = result$categories,
  point_color = "type",
  point_shape = "ownership",
  point_size = "capacity",
  point_colors = c(
    "Hospital" = "#d73027",
    "Clinic" = "#fc8d59",
    "Health Center" = "#fee08b"
  ),
  point_shapes = c("Public" = 16, "Private" = 17),
  label = "Health Facilities"
)

plot(advanced_map)
```

## Custom Bar Charts

```{r custom-bar-chart, eval = FALSE}
# Create bar chart with custom label colors
custom_bar <- create_bar_plot(
  summary_dt = result$national_summary,
  categories = result$categories,
  show_labels = TRUE,
  color_labels = "darkblue"
)

plot(custom_bar)
```

# Individual Function Workflow

For maximum control, use individual functions:

```{r individual-workflow, eval = FALSE}
# Step 1: Load and validate data
kenya_data <- load_default_data()
points <- validate_points(health_facilities)
cats <- default_categories()

# Step 2: Compute accessibility surface
travel_results <- compute_travel_cost(
  points, 
  kenya_data$transitionMatrix, 
  kenya_data$populationRaster
)

# Step 3: Process data
access_data <- process_access_data(
  travel_results$access, 
  travel_results$pop, 
  cats
)

# Step 4: Generate summaries
summary_stats <- generate_summary(access_data, cats)

# Step 5: Create custom visualizations
map_plot <- create_map(
  access_data, points, kenya_data$adm0, kenya_data$adm1, cats,
  point_color = "blue", 
  point_size = 6,
  show_scale = FALSE
)

bar_plot <- create_bar_plot(
  summary_stats, cats, 
  color_labels = "red"
)

combined_plot <- combine_plots(
  map_plot, bar_plot,
  map_width = 0.8  # Give map more space
)

# Step 6: Save with custom options
save_outputs(
  plots = list(
    map_plot = map_plot,
    bar_plot = bar_plot,
    combined_plot = combined_plot
  ),
  summary_dt = summary_stats,
  path = "./custom_outputs",
  name = "custom_analysis",
  save_opts = list(dpi = 300, width = 2, height = 2),
  formats = c("png", "pdf")
)
```

# Working with Your Own Data

## Coordinate Validation

The package automatically detects coordinate columns:

```{r coordinate-validation}
# Works with different column names
coords_df1 <- data.frame(longitude = 36.8, latitude = -1.3)
coords_df2 <- data.frame(lon = 36.8, lat = -1.3)
coords_df3 <- data.frame(x = 36.8, y = -1.3)

# All work automatically
points1 <- validate_points(coords_df1)
points2 <- validate_points(coords_df2)
points3 <- validate_points(coords_df3)

# Check results
head(points1)
```

## Using Your Own Spatial Data

```{r own-data, eval = FALSE}
# For other countries/regions
result <- accessibility(
  points = my_facility_coordinates,
  transitionMatrix = my_friction_surface,
  populationRaster = my_population_data,
  adm0 = my_country_boundary,
  adm1 = my_admin_boundaries,
  categories = my_custom_categories,
  label = "My Facilities",
  path = "./my_outputs",
  name = "my_analysis"
)
```

# Color Palettes

## Using Different Palettes

```{r palette-examples, eval = FALSE}
# Viridis palette
viridis_cats <- create_categories(
  breaks = c(0, 30, 60, 90, 120, Inf),
  palette = "viridis"
)

# ColorBrewer palette
brewer_cats <- create_categories(
  breaks = c(0, 30, 60, 90, 120, Inf),
  palette = "RdYlBu",
  direction = -1  # Reverse colors
)

# Custom colors
custom_cats <- create_categories(
  breaks = c(0, 30, 60, 90, 120, Inf),
  labels = c("Excellent", "Good", "Fair", "Poor", "Very Poor"),
  colors = c("#2166ac", "#762a83", "#af8dc3", "#e7d4e8", "#f7f7f7")
)
```

